Tree establishment in models of forest dynamics: a key priority for further research

### 2023-02-21 manuscript/manuscriptRW.Rmd
comments and changes until discussion

Two models, ForClim and ForCEEPs, were used here with two alternative
variants. ForClim variant 1 [@bugmann1996] is based on a recruitment
module that adheres closely to the concept introduced by @botkin1972,
whereas ForClim variant 11 is adopting a rather different approach that
considers space filling along with different 'points of entry' for the
action of specific environmental factors [@huber2020]. These two model
variants allow us to evaluate a more process-based module (variant 11)
against a simple module (variant 1), while the rest of the model
structure is identical. Similarly, the two ForCEEPS variants allow us to
isolate the importance of the canopy feedback (i.e. simulated actual
composition and relative abundance of species in the plot)via seed trees
for the quantity and quality of simulated ingrowth, as one variant
includes this feedback whereas the other does not.

@@

The simulations were run for mixed species forests (not multiple
single-species simulations) using mixtures of eleven species or genera
for which ingrowth data of sufficient quality were available from
EuFoRIa: *Fagus sylvatica, Picea abies, Abies alba, Carpinus betulus,
Tilia cordata, Acer pseudoplatanus, Betula spp., Fraxinus excelsior,
Quercus spp., Alnus glutinosa, Pinus sylvestris.* The same set of
species were used in all the 200 sites. Two models included their
standard set of species for the simulations which is much larger (i.e.,
ForClim 1, ForClim 11, TreeMig). In three models, fewer than these
eleven species were simulated (4C, xComp, LPJ-GUESS) (Table
\@ref(tab:tableModels)). In 4C onl*y Fagus sylvatica*, *Picea abies*,
*Betula* spp., *Quercus* spp. and *Pinus sylvestris* are parameterised.
XComp simulations did not consider *Fraxinus excelsior* due to a
depreciated species parameterisation. In LPJ-GUESS, *Acer
pseudoplatanus* and *Alnus glutinosa* are not parameterized and
therefore these species could not be included in the simulation.
Finally, the model aDGVM2 does not simulate individual species. Rather,
community assembly processes and trait filtering generate plant
communities that are adjusted to the biotic and abiotic conditions and
plants can be classified into different ecological strategies based on
their trait values in a post-processing step.

@@

The simulation results were analyzed regarding the ingrowth levels
(i.e., ingrowth number per unit time and space), species composition of
the ingrowth, and ingrowth mortality. The observed data were modeled
using a Generalized Additive Model [@wood2011] with a negative binomial
distribution and restricted maximum likelihood to better understand the
trend and levels of the observed ingrowth, relative to the simulation
results per model.

Ingrowth levels were evaluated along three key environmental gradients,
i.e., stand basal area (the basal area of all the trees in each sample)
as a proxy of light availability at the forest floor, the annual
degree-day sum (DD) [@allen1976; @fischlin1995] as a proxy for growing
season warmth, and the climatic water balance (CWB) as a proxy for soil
moisture [@speich2019]. The mean ingrowth values were calculated as the
mean across the 200 samples per site.

We evaluated the species composition across models and in relation to
the observed data by calculating the Shannon index based on the relative
proportion of the species in terms of basal area. The higher the value
of the index, the higher the species diversity at a particular site and
sample (Equation \ref{eqpiShannon}).

```{=tex}
\begin{equation}
H_n = - \sum_{i = 1}^{s}{p_{n,i}ln(p_{n,i})}\label{eqpiShannon}
\end{equation}
```
```{=tex}
\begin{equation}
p_{n,i} = \frac{r_{BA_{n,i}}}{Total_{BA_{n}}} \label{eqpiShannon2}
\end{equation}
```
where $s$ is the total number of species simulated that have ingrowth
basal area above zero in sample $n$; $p_{n,i}$ is the proportion of the
species $i$ calculated as the ingrowth basal area ($r_{BA_{n,i}}$) for
that species $i$ relative to the stand basal area ($Total_{BA_{n}}$) of
the sample $n$.

@@

```{=tex}
\begin{equation}
\bar{R}_{BA_{share_{t,i}}} = \frac{\bar{r}_{t,i}}{\sum_{i = 1}^{s}{\bar{r}_{t,i}}} \label{eq:BAshare}
\end{equation}
```
where $s$ is the total number of species simulated; $\bar{r}_{t,i}$ is
the mean ingrowth of the species $i$ per site $t$ across the available
samples at that site.

The data from 35 sites that had a diameter threshold of 10 cm only, and
therefore ingrowth rates for these sites at 7 cm are not known. The 35
sites were removed from the simulated data to make a fair comparison
regarding the ingrowth of the five main species. Furthermore, regarding
the ingrowth basal area share per species ($\bar{R}_{BA_{share_{i}}}$),
we categorized this as zero when the total ingrowth basal area (
$\sum_{i = 1}^{s}{\bar{r}_{t,i}}$ ) equaled zero, and also when the
ingrowth basal area of that species ($\bar{r}_{t,i}$) was zero.

Mortality in tree establishment was assessed based on the ratio of
ingrowth between the 7 and 10 cm diameter thresholds. We used the
Reineke self-thinning rule [@reineke1933; @pretzsch2003] as a reference
to estimate whether the different ingrowth values were above or below
the expected theoretical rate between these two diameters. The Reineke
self-thinning rule is usually calculated from even-aged, single species
stands and assumes a relationship between the number of stems and the
quadratic mean diameter in fully stocked pure stands. The value used in
our comparisons was 1.77, calculated using Equation \ref{eqReineke}.

@@

The ingrowth levels varied strongly across the 15 models and typically
did not match the levels found in the forest reserve data (Figure
\@ref(fig:over)). Ingrowth was overestimated in most models for both the
7 and 10 cm diameter thresholds, with the exception of the empirical
stand model SIBYLA, the landscape model Landis-II and the global model
aDGVM2, which estimated ingrowth rates at the lower end of the
plausibility interval of the observed data (Figure \@ref(fig:over)). The
models with the largest overestimation were the stand model PICUS and
the landscape model TreeMig. For most models, the variability of
simulated ingrowth levels across the 200 sites (visible from the
interquartile range in the box plots of Figure \@ref(fig:over)) was
similar to or smaller than observed, with the notable exception of
PICUS, where simulated ingrowth variability was much larger than that of
any other model.

```{r over, echo=FALSE, fig.cap="Mean ingrowth rates across all samples per site, plotted for the 200 sites and for each model. The red dotted lines shows the 2.5 and 97.5 percentiles for the 7 cm diameter threshold in the observed data.", out.width='100%'}
        knitr::include_graphics(here::here("figures", "recOverAll710.png"))
```

When evaluating the representation of ingrowth rates along key gradients
of light availability (basal area), temperature (degree-days) and soil
moisture (climatic water balance), distinct features emerge: the models
reproduced the effect of basal area (Figure \@ref(fig:nvTrend710BA))
much better than the effects of the climatic gradients, where they
featured varying effects (Figures S1 and S2).

In the observed data, ingrowth rates decreased clearly with increasing
stand basal area (Figure \@ref(fig:nvTrend710BA)). This trend was
captured by the Landis-II model only, albeit at much lower values of
basal area than in the observations. The other models featured
distinctly different trends, such as (1) an increase in ingrowth with
increasing total basal area followed by the absence of ingrowth at high
values of basal area (ForCEEPS, ForCEEPS(f), FORMIND, TreeMig and
LPJ-GUESS), (2) almost constant ingrowth levels with basal area (ForClim
11, xComp, LandClim and aDGVM2), or (3) an increase of ingrowth up to a
certain value of stand basal area followed by a decrease at even higher
values of stand basal area, with model-specific thresholds (4C, ForClim
1, PICUS, SIBYLA and iLand).

The models did not match the observed ingrowth patterns across gradients
of soil moisture and temperature (Figures S1 and S2), and neither was
there a clear pattern across models. Regarding the soil moisture
gradient (Figure S1), a group of models featured decreasing ingrowth
with increasing soil moisture (4C, ForClim 1, ForClim11, SIBYLA, xComp,
PICUS and LandClim), while a few models showed the opposite trend
(ForCEEPs(f), FORMIND and LPJ-GUESS). Across the temperature gradient
(Figure S2), all the models that featured a decrease in ingrowth rates
with increasing soil moisture showed the opposite trend (i.e., more
ingrowth with increasing temperature), except for 4C. The other models
did not feature a clear pattern.

Both observations and simulated data had no ingrowth in some samples and
at some sites (Table S1). The observed data had slightly more than 3% of
the samples with no ingrowth. Only three models had a larger proportion
of no ingrowth (4C, Landis-II and aDVM2). Three models always simulated
ingrowth for both the 7 and 10 cm threshold (TreeMig and LPJ-GUESS),
i.e., they did not feature any zero values. The other models had a very
low percentage of samples with no ingrowth (0.01% - 1.31%), i.e., they
had distinctly fewer occurrences of zeros compared to the observations.
There was no significant relationship between the mean complexity of the
establishment module (Table \@ref(tab:tableModels)) and the
overestimation proportion at a diameter threshold of 7 cm (SM3).

@@

Most models matched the level of diversity of the observed data quite
well (Figure \@ref(fig:H710)), although they consistently underestimated
its variability across sites. Four models overestimated ingrowth
diversity: 4C, PICUS, Landis-II and LPJ-GUESS, the latter mostly for the
10 cm diameter threshold. The model 4C is a special case as it simulated
five species only, i.e., its diversity values are not directly
comparable to those of the other models, nor to the observations. Only
one model, SIBYLA, consistently underestimated ingrowth diversity.

In most models, there were only small but significant differences in the
species diversity of ingrowth between 7 and 10 cm across sites (Tables
S2-S4). Furthermore, three models (4C, SIBYLA, Landis-II) maintained the
ingrowth diversity between the 7 and 10 cm thresholds (i.e., the
differences between them were not significant, see Table S3), and the
same was evident from the observed data. There were a few models with a
strong increase of ingrowth diversity between 7 and 10 cm (ForCEEPS,
TreeMig, LPJ-GUESS, Figure \@ref(fig:H710)).

@@

For both the observed and the simulated data, and both ingrowth
thresholds, species diversity was lower in the ingrowth compared to
stand-level diversity (Figures S3 and S4). Yet, models simulating higher
diversity at the stand level featured higher diversity in ingrowth
(Figure S3 and S4). Most models captured reasonably well or
overpredicted species diversity at the stand level (Figure
\@ref(fig:combinedRichness)A, center; cf. Figure S5), but a
characteristic feature was that simulations had a much lower variability
of diversity compared to observations. Models overpredicting species
richness included ForClim1, ForClim 11, FORMIND and TreeMig (Figure
\@ref(fig:combinedRichness)A, left). Several models underpredicted
species diversity, i.e., LPJ-GUESS, ForCEEPS(f) and 4C (Figure
\@ref(fig:combinedRichness)A, right). The majority of the models
captured reasonably well the ingrowth diversity at lower values of
diversity, but most of them strongly underestimated ingrowth diversity
for higher observed diversity levels (Figure
\@ref(fig:combinedRichness)B, cf. Figure S6). Both the models with
feedback and those without feedback from the adult trees to ingrowth
(via seed production) underestimated ingrowth species diversity compared
to the observations. Again, simulated diversity was considerably less
variable than observed diversity (Fig. \@ref(fig:combinedRichness)B).

@@

First, some models overestimated ingrowth across the entire gradients
for at least one of the five main tree species (Figure \@ref(fig:diff),
e.g., ForCEEPS, ForCEEPS(f), LandClim or LPJ-GUESS), while other models
overestimated ingrowth for more than one species for a large part of the
climate gradients, such as xComp, PICUS, SIBYLA or LPJ-GUESS (cf. red
colors in Figure \@ref(fig:diff)). These trends were consistent for both
ingrowth thresholds, although the exact changes in the share of the
ingrowth basal area were somewhat different (Figure S7).

Second, there was some congruence in the simulation results by species
across the models. The share in the ingrowth by Silver fir (*Abies
alba*) was captured well across the gradient by most models except for
xComp, PICUS, LandClim and Landis-II, which overestimated the ingrowth
share of this species. There were also some peculiarities evident for
some models, such as 4C that did not simulate *Abies alba*. *Fagus
sylvatica* ingrowth was underestimated by many models across the
environmental gradients, including ForCEEPS(f) and FORMIND, which
featured negligible ingrowth at most sites for this species. Most of the
models overestimated the share of *Picea abies* ingrowth in the cold-wet
part of the gradients. *Pinus sylvestris* and *Quercus* spp. represented
a smaller share of the ingrowth in the observed data mostly at warm-dry
sites. Many models, however, erroneously featured ingrowth for these two
species along most of the climatic gradients although at a low basal
area share. Five models (4C, FORMIND, SIBYLA, xComp, and LandClim) had
almost no ingrowth of *Pinus sylvestris* and *Quercus* spp., or did not
feature any ingrowth of these species at all.

@@

The fact that some models featured lower mortality than expected by
Reineke's self-thinning rule may point to facilitation, or simply a
higher mortality rate before the trees had reached 7 cm diameter. Yet,
the case of models such as xComp, PICUS and TreeMig, whose mortality was
well above the self-thinning line, indicates that they feature higher
ingrowth mortality in mixed-species forests compared to even-aged,
single-species forests. Some models yielded a mortality rate of ingrowth
that is broadly compatible with the self-thinning rule (i.e., 4C and
LandClim).

The three models with the largest overestimation of the proportion of
ingrowth at the 7 cm threshold featured the highest mortality between
the 7 and 10 cm thresholds (Figure \@ref(fig:mortality13)A). The other
models that overestimated ingrowth had mortalities lower than the
expected self-thinning ratio (Figure \@ref(fig:mortality13)B). It is
noteworthy that most of the models that featured a low ratio of ingrowth
between 7 and 10 cm (i.e., ForCEEPS, ForClim 1, ForClim 11 and
LandClim) (Figure S8) had only a small overestimation of ingrowth at the
7 cm threshold (Figure \@ref(fig:mortality11) cf. Figure S8). aDGVM2 was
the only model that underestimated ingrowth at the 7 cm threshold while
featuring a low mortality between the two thresholds (Figure
\@ref(fig:mortality13)C), and only a few models were close to the
correct ingrowth rate at 7 cm while also being close to the Reineke line
(SIBYLA, Landis-II; cf. Figure \@ref(fig:mortality13)D; and 4C as well
as LandClim).

Lastly, some models that started at a low number of individuals (7 cm
threshold) had a lower decrease of ingrowth between 7 and 10 cm dbh
(ratio 7-10 cm) (Figure S10). In these models, competition and
self-thinning are either not pronounced or must have occurred before the
species had reached 7 cm. However, this pattern was not consistent
across models (Figure S9). It is noteworthy that there is a relationship
between the ingrowth rate at 7 cm and the mortality rate: most of the
models showed a positive effect (higher ingrowth at 7 cm is coupled to
an increase of the mortality between 7 and 10 cm). While some models
showed a negative effect, this linear trend was not pronounced and not
always significant regarding the slope (Figure S9 and Table S2).

@@

Tree establishment is a fundamental process in forest dynamics, in
particular with increasing disturbance under climate change. Correctly
capturing tree ingrowth in dynamic models is fundamental to, for
example, evaluate post-disturbance dynamics as it will define forest
state in the first decades [@seidl2022]. If this initial phase is not
captured well, we can not properly assess aspects such as forest
resilience or the timing, magnitude and progression of carbon
sequestration. In this study, for the first time, projections of tree
establishment from multiple models of forest dynamics were confronted
with a unique dataset from unmanaged forest reserves across a large
environmental gradient in Europe.

The observed data that we used [@käber2022] are also exceptional,
particularly with respect to the number of events captured, which is
essential for a highly 'noisy' process such as tree establishment. The
use of this dataset for model benchmarking provided novel insights on
the capability of state-of-the art models to simulate ingrowth levels,
species composition, and mortality in an early stage of tree life, i.e.,
between a tree diameter of 7 and 10 cm. Overall, by adopting this
approach a much broader understanding resulted than if we had used these
data for model calibration: it is primarily from the failure of the
models that we can gain ecological insights [e.g., @trugman2022].

@@

Most of the models tended to overestimate tree ingrowth rates. This has
potentially far-reaching implications, e.g., regarding biomass (and thus
carbon) turnover, with a potential overestimation of the capacity for
forest carbon sequestration [@pugh2019; @pan2011]. Yet, we focused on
one specific stage, i.e., ingrowth into the 7 and 10 cm diameter
classes. Trees at this size contribute little to carbon sequestration,
and if the excess ingrowth rate at this stage is compensated -- in the
models -- soon thereafter by higher mortality, overall simulated
ingrowth may still yield trustworthy results at the level of stand
structure and related forest attributes such as basal area or biomass.

From a modeling point of view, excess ingrowth inevitable requires
excess mortality rates in a later stage, either -- as observed for some
models -- between the 7 and 10 cm diameter thresholds, before the 7 cm
are reached, or soon after the 10 cm threshold has been crossed. In any
case, correcting at early stages for the expected forest densities at
later stages is equivalent to compensating for a first error (excess
ingrowth) by a second error (excess mortality). Thus, many models appear
to get the 'right answer', i.e., correct stand structure and
composition, for the wrong reasons. It is highly likely that biased
projections will result, because the two errors are unlikely to be
perfectly linked and thus will not always compensate each other. Hence,
this structural problem of most models investigated here is problematic
particularly if the models are to be used under novel conditions such as
under climate change [e.g., @huber2021] or in a decision support context
[e.g., @thrippleton2021].

>>>>

Two models, ForClim and ForCEEPS, were used here with two alternative
variants. ForClim variant 1 [@bugmann1996] is based on a recruitment
module that adheres closely to the concept introduced by @botkin1972,
whereas ForClim variant 11 is adopting a rather different approach that
considers space filling along with different 'points of entry' for the
action of specific environmental factors [@huber2020]. These two model
variants allow us to evaluate a more process-based module (variant 11)
against a simple module (variant 1), while the rest of the model
structure is identical. Similarly, the two ForCEEPS variants allow us to
isolate the importance of the canopy feedback (i.e. simulated actual
composition and relative abundance of species in the plot)via seed trees
for the quantity and quality of simulated ingrowth, as one variant
includes this feedback whereas the other does not.

@@

The simulations were run for mixed species forests (not multiple
single-species simulations) using mixtures of eleven species or genera
for which ingrowth data of sufficient quality were available from
EuFoRIa: *Fagus sylvatica, Picea abies, Abies alba, Carpinus betulus,
Tilia cordata, Acer pseudoplatanus, Betula spp., Fraxinus excelsior,
Quercus spp., Alnus glutinosa, Pinus sylvestris.* The same set of
species were used in all the 200 sites. Two models included their
standard set of species for the simulations which is much larger (i.e.,
ForClim 1, ForClim 11, TreeMig). In three models, fewer than these
eleven species were simulated (4C, xComp, LPJ-GUESS) (Table
\@ref(tab:tableModels)). In 4C only *Fagus sylvatica*, *Picea abies*,
*Betula* spp., *Quercus* spp. and *Pinus sylvestris* are parameterised.
XComp simulations did not consider *Fraxinus excelsior* due to a
depreciated species parameterisation. In LPJ-GUESS, *Acer
pseudoplatanus* and *Alnus glutinosa* are not parameterized and
therefore these species could not be included in the simulation.
Finally, the model aDGVM2 does not simulate individual species. Rather,
community assembly processes and trait filtering generate plant
communities that are adjusted to the biotic and abiotic conditions and
plants can be classified into different ecological strategies based on
their trait values in a post-processing step.

@@

The simulation results were analyzed regarding the ingrowth levels
(i.e., ingrowth number per unit time and space), species composition of
the ingrowth, and ingrowth mortality. The observed data were modeled
using a Generalized Additive Model [@wood2011] with a negative binomial
distribution and restricted maximum likelihood to better understand the
environmental gradients trend in the levels of the observed ingrowth,
relative to the simulation results per model.

Ingrowth levels were evaluated along three key environmental gradients,
i.e., stand basal area (the basal area of all the trees in each sample
including tree establishment) as a proxy of light availability at the
forest floor, the annual degree-day sum (DD) [@allen1976; @fischlin1995]
as a proxy for growing season warmth, and the climatic water balance
(CWB) as a proxy for soil moisture [@speich2019]. The mean ingrowth
values were calculated as the mean across the 200 samples per site.

We evaluated the species diversity across models and in relation to the
observed data by calculating the Shannon index based on the relative
proportion of the species in terms of basal area. The Shannon index was
calculated for the establishment ($H_{n,r}$) (Equation
\ref{eqpiShannonR}) and at the stand level ($H_{n,s}$)(Equation
\ref{eqpiShannon}). The higher the value of the index, the higher the
species diversity at a particular site and sample.

We also tested for significant trends across XXXXX !!!!!

```{=tex}
\begin{equation}
H_{n,r} = - \sum_{i = 1}^{s}{p_{n,i}ln(p_{n,r,i})}\label{eqpiShannonR}
\end{equation}
```
```{=tex}
\begin{equation}
p_{n,r,i} = \frac{r_{BA_{n,i}}}{TotalR_{BA_{n}}} \label{eqpiShannonR2}
\end{equation}
```
```{=tex}
\begin{equation}
H_{n,s} = - \sum_{i = 1}^{s}{p_{n,i}ln(p_{n,i})}\label{eqpiShannon}
\end{equation}
```
```{=tex}
\begin{equation}
p_{n,s,i} = \frac{BA_{n,i}}{Total_{BA_{n}}} \label{eqpiShannon2}
\end{equation}
```
where $s$ is the total number of species present and have basal area
above zero in sample $n$; $p_{n,r,i}$ is the proportion of the species
$i$ calculated as the ingrowth basal area ($r_{BA_{n,i}}$) for that
species $i$ relative to the recruited basal area ($TotalR_{BA_{n}}$) of
the sample $n$; $p_{n,s,i}$ is the proportion of the species $i$
calculated as the basal area ($BA_{n,i}$) for that species $i$ relative
to the total basal area ($TotalR_{BA_{n}}$) of the sample $n$.

@@

```{=tex}
\begin{equation}
\bar{R}_{BA_{share_{t,i}}} = \frac{\bar{r}_{BA_{t,i}}}{\sum_{i = 1}^{s}{\bar{r}_{{BA_{t,i}}}}} \label{eq:BAshare}
\end{equation}
```
where $s$ is the total number of species simulated; $\bar{r}_{t,i}$ is
the mean basal area ingrowth of the species $i$ per site $t$ across the
available samples at that site.

We removed 35 sites from the comparisons at 7 cm ingrowth, because field
data did not include these measurements. Furthermore, regarding the
ingrowth basal area share per species ($\bar{R}_{BA_{share_{i}}}$), we
categorized this as zero when the ingrowth basal area of that species
($\bar{r}_{BA_{t,i}}$) was zero but also when both the ingrowth basal
area of that species ($\bar{r}_{BA_{t,i}}$) and the total ingrowth basal
area ( $\sum_{i = 1}^{s}{\bar{r}_{BA_{t,i}}}$ ) equaled zero.

Mortality in tree establishment was assessed based on the ratio of
ingrowth between the 7 and 10 cm diameter thresholds. We used the
Reineke self-thinning rule [@reineke1933; @pretzsch2003] as a reference
to estimate whether the different ingrowth values were above or below
the expected theoretical rate between these two diameters. The Reineke
self-thinning rule is usually calculated from even-aged, single species
stands and assumes a relationship between the number of stems and the
quadratic mean diameter in fully stocked pure stands. The value used in
our comparisons was 1.77 (i.e., we expect stem numbers at 7cm to be 77%
higher than at 10cm), calculated using Equation \ref{eqReineke}.

@@

The ingrowth levels varied strongly across the 15 models and typically
did not match the levels found in the forest reserve data (Figure
\@ref(fig:over)). Ingrowth was overestimated in most models for both the
7 and 10 cm diameter thresholds, with the exception of the empirical
stand model SIBYLA, the landscape model Landis-II and the global model
aDGVM2, which estimated ingrowth rates at the lower end of the
plausibility interval of the observed data (Figure \@ref(fig:over)). The
models with the largest overestimation were the stand model PICUS and
the landscape model TreeMig. For most models, the variability of
simulated ingrowth levels across the 200 sites (visible from the
interquartile range in the box plots of Figure \@ref(fig:over)) was
similar to or smaller than observed, with the notable exception of
PICUS, where simulated ingrowth variability was much larger.

```{r over, echo=FALSE, fig.cap="Mean ingrowth rates across all samples per site, plotted for the 200 sites and for each model. The red dotted lines shows the 2.5 and 97.5 percentiles for the 7 cm diameter threshold in the observed data. There are two boxplots for each model where the left lighted boxplot corresponds to 7cm and the right darker boxplot to 10cm", out.width='100%'}
        knitr::include_graphics(here::here("figures", "recOverAll710.png"))
```

When evaluating the representation of ingrowth levels along key
gradients of light availability (basal area), temperature (degree-days)
and soil moisture (climatic water balance), distinct features emerge:
the models reproduced the effect of basal area in magnitude and patterns
(Figure \@ref(fig:nvTrend710BA)) better than the effects of the climatic
gradients, where they featured varying effects (Figures S1 and S2).

In the observed data, ingrowth levels decreased clearly with increasing
stand basal area (Figure \@ref(fig:nvTrend710BA)). This trend was
captured by the Landis-II model only, albeit at much lower values of
basal area than in the observations. The other models featured
distinctly different trends, such as (1) an increase in ingrowth levels
with increasing total basal area followed by the absence of ingrowth at
high values of basal area (ForCEEPS, ForCEEPS(f), FORMIND, TreeMig and
LPJ-GUESS), (2) almost constant ingrowth levels with basal area (ForClim
11, xComp, LandClim and aDGVM2), or (3) an increase of ingrowth up to a
certain value of stand basal area followed by a decrease at even higher
values of stand basal area, with model-specific thresholds (4C, ForClim
1, PICUS, SIBYLA and iLand).

The models did not match the observed ingrowth patterns across gradients
of soil moisture and temperature (Figures S1 and S2), and neither was
there a clear pattern across models. Regarding the soil moisture
gradient (Figure S1), a group of models featured decreasing ingrowth
with increasing soil moisture (4C, ForClim 1, ForClim11, SIBYLA, xComp,
PICUS, LandClim, TreeMig and iLand), while a few models showed the
opposite trend (ForCEEPS(f), FORMIND and LPJ-GUESS). Across the
temperature gradient (Figure S2), all the models that featured a
decrease in ingrowth rates with increasing soil moisture showed the
opposite trend (i.e., more ingrowth with increasing temperature), except
for 4C. The other models did not feature a clear pattern.

Both observations and simulated data had no ingrowth in some samples and
at some sites (Table S1). The observed data had 3% of the samples with
no ingrowth. Only three models had a larger proportion of no ingrowth
(4C, Landis-II and aDVM2). CHECK! No LPJ-GUESS anymore with the new data
submission!!!! Two models always simulated ingrowth for both the 7 and
10 cm threshold (TreeMig and LPJ-GUESS), i.e., they did not feature any
zero values. The other models had a very low percentage of samples with
no ingrowth (0.01% - 1.31%), i.e., they had distinctly fewer occurrences
of zeros compared to the observations. There was no significant
relationship between the mean complexity of the establishment module
(Table \@ref(tab:tableModels)) and the overestimation proportion at a
diameter threshold of 7 cm (SM3).

@@

Most models matched the level of diversity of the observed data quite
well as quantified by the Shannon index (Figure \@ref(fig:H710)),
although they consistently underestimated its variability across sites.
Four models overestimated ingrowth diversity: 4C, PICUS, Landis-II and
LPJ-GUESS, the latter mostly for the 10 cm diameter threshold. The model
4C is a special case, as it simulated five species only, i.e., its
diversity values are not directly comparable to those of the other
models, nor to the observations. Only one model, SIBYLA, consistently
underestimated ingrowth diversity.

In most models, there were only small but significant differences in the
species diversity of ingrowth between 7 and 10 cm across sites (Tables
S2-S4). Three models (4C, SIBYLA, Landis-II) maintained the ingrowth
diversity between the 7 and 10 cm thresholds (i.e., the differences
between them were not significant, see Table S3), and the same was
evident from the observed data. There were a few models with a strong
increase of ingrowth diversity between 7 and 10 cm (ForCEEPS, TreeMig,
LPJ-GUESS, FORMIND, Figure \@ref(fig:H710)).

@@

For both the observed and the simulated data, and both ingrowth
thresholds, species diversity was lower in the ingrowth compared to
stand-level diversity (Figures S3 and S4). Yet, models simulating higher
diversity at the stand level featured higher diversity in ingrowth
(Figure S3 and S4). Most models captured reasonably well or
overpredicted species diversity at the stand level (Figure
\@ref(fig:combinedRichness)A, center; cf. Figure S5), but a
characteristic feature was that simulations had a much lower variability
of diversity compared to observations. Models overpredicting species
diversity included ForClim1, ForClim 11, FORMIND and TreeMig (Figure
\@ref(fig:combinedRichness)A, left). Several models underpredicted
species diversity, i.e., LPJ-GUESS, ForCEEPS(f) and 4C (Figure
\@ref(fig:combinedRichness)A, right). The majority of the models
captured reasonably well the ingrowth diversity at lower values of
diversity, but most of them strongly underestimated ingrowth diversity
for higher observed diversity levels (Figure
\@ref(fig:combinedRichness)B, cf. Figure S6). Both the models with
feedback and those without feedback from the adult trees to ingrowth
(via seed production) underestimated ingrowth species diversity compared
to the observations. Again, simulated diversity was considerably less
variable than observed diversity (Fig. \@ref(fig:combinedRichness)B).

@@

First, some models overestimated ingrowth across the entire gradients
for at least one of the five main tree species (Figure \@ref(fig:diff),
e.g., ForCEEPS, ForCEEPS(f), LandClim or LPJ-GUESS), while other models
overestimated ingrowth for more than one species for a large part of the
climate gradients, such as xComp, PICUS, SIBYLA or LPJ-GUESS (cf. red
colors in Figure \@ref(fig:diff)). These trends were consistent for both
ingrowth thresholds, although the exact changes in the share of the
ingrowth basal area were different (Figure S7).

Second, there was some congruence in the simulation results by species
across the models. The share in the ingrowth of Silver fir (*Abies
alba*) was captured well across the gradient by most models except for
xComp, PICUS, LandClim and Landis-II, which overestimated the ingrowth
share of this species. There were also some peculiarities evident for
some models, such as 4C that did not simulate *Abies alba*. The patterns
across models for *Fagus sylvatica* ingrowth were more complex, the
ingrowth was underestimated by many models across the environmental
gradients featuring negligible ingrowth at most sites, while others
overestimate it in different parts of the environmental space. Most of
the models overestimated the share of *Picea abies* ingrowth in the
cold-wet part of the gradients. *Pinus sylvestris* and *Quercus* spp.
represented a smaller share of the ingrowth in the observed data mostly
at warm-dry sites. Many models, however, erroneously featured ingrowth
for these two species along most of the climatic gradients although at a
low basal area share. Five models (4C, FORMIND, SIBYLA, xComp, and
LandClim) had almost no ingrowth of *Pinus sylvestris* and *Quercus*
spp., or did not feature any ingrowth of these species at all, close to
the patterns from the observed data.

@@

The fact that some models (and observed data) featured lower mortality
than expected by Reineke's self-thinning rule may point to facilitation,
or simply a higher mortality rate before the trees had reached 7 cm
diameter. Yet, the case of models such as xComp, PICUS and TreeMig,
whose mortality was well above the self-thinning line, indicates that
they feature higher ingrowth mortality in mixed-species forests compared
to even-aged, single-species forests. Some models yielded a mortality
rate of ingrowth that is broadly compatible with the self-thinning rule
(e.g., 4C and LandClim).

The three models with the largest overestimation of the proportion of
ingrowth at the 7 cm threshold featured the highest mortality between
the 7 and 10 cm thresholds (Figure \@ref(fig:mortality13)A). The other
models that overestimated ingrowth had mortalities lower than the
expected self-thinning ratio with the exception of LandClim (Figure
\@ref(fig:mortality13)B). It is noteworthy that most of the models that
featured a low ratio of ingrowth between 7 and 10 cm (i.e., ForCEEPS,
ForClim 1, ForClim 11 and iLand) had only a small overestimation of
ingrowth at the 7 cm threshold (Figure \@ref(fig:mortality11)). aDGVM2
was the only model that underestimated ingrowth at the 7 cm threshold
while featuring a low mortality between the two thresholds (Figure
\@ref(fig:mortality13)C), and only a few models were close to the
correct ingrowth rate at 7 cm while also being close to the Reineke line
(SIBYLA, Landis-II; cf. Figure \@ref(fig:mortality13)D; and 4C as well
as LandClim).

Lastly, some models that had a low number of individuals at 7 cm
threshold also had a smaller decrease of ingrowth between 7 and 10 cm
(ratio 7-10 cm) (Figure S10). In these models, competition and
self-thinning are either not pronounced or must have occurred before the
species had reached 7 cm. However, this pattern was not consistent
across models (Figure S10). It is noteworthy that there is a
relationship between the ingrowth rate at 7 cm and the mortality rate:
most of the models showed a positive effect (higher ingrowth at 7 cm is
coupled to an increase of the mortality between 7 and 10 cm). While some
models showed a negative effect, this linear trend was not pronounced
and not always significant regarding the slope (Figure S9 and Table S6).

@@

Tree establishment is a fundamental process in forest dynamics, in
particular under changing conditions such as those caused by climate
change. Correctly capturing tree ingrowth in dynamic models is
fundamental to, for example, evaluate post-disturbance dynamics as it
will define forest state in the first decades [@seidl2022]. If this
initial phase is not captured well, we can not properly assess aspects
such as forest resilience or the timing, magnitude and progression of
carbon sequestration. In this study, for the first time, projections of
tree establishment from multiple models of forest dynamics were
confronted with a unique dataset from unmanaged forest reserves across a
large environmental gradient in Europe.

The observed data that we used [@käber2022] are exceptional,
particularly with respect to the number of events captured (number of
sites and repetitive measurements), which is essential for a highly
'noisy' process such as tree establishment. The use of this dataset for
model benchmarking provided novel insights on the capability of
state-of-the-art models to simulate ingrowth levels, species
composition, and mortality in an early stage of tree life, i.e., between
a tree diameter of 7 and 10 cm. Overall, by adopting this approach a
much broader understanding resulted than if we had used these data for
model calibration: it is primarily from the shortcomings of the models
that we can gain ecological insights [e.g., @trugman2022].

@@

Most of the models overestimated tree ingrowth rates. This has
potentially far-reaching implications, e.g., regarding biomass (and thus
carbon) turnover, with a potential overestimation of the capacity for
forest carbon sequestration [@pugh2019; @pan2011]. Yet, we focused on
one specific stage, i.e., ingrowth into the 7 and 10 cm diameter
classes. Trees at this size contribute little to carbon sequestration,
and if the excess ingrowth rate at this stage is compensated -- in the
models -- soon thereafter by higher mortality, overall simulated
ingrowth may still yield trustworthy results at the level of stand
structure and related forest attributes such as basal area or biomass.

From a modeling point of view, excess ingrowth inevitably requires
excess mortality rates in a later stage, either -- as observed for some
models -- between the 7 and 10 cm diameter thresholds, or soon after the
10 cm threshold has been crossed. In any case, correcting at early
stages for the expected forest densities at later stages is equivalent
to compensating for a first error (excess ingrowth) by a second "error"
(excess mortality). Thus, many models appear to get the 'right answer',
i.e., correct stand structure and composition, for the wrong reasons. It
is highly likely that biased projections will result, because the two
errors are unlikely to be perfectly linked and thus will not always
compensate each other. Hence, this structural problem of most models
investigated here is problematic particularly if the models are to be
used under novel conditions such as under climate change [e.g.,
@huber2021] or in a decision support context [e.g., @thrippleton2021].
